# Data structure and prevalence

Microbiome count data is structurally sparse and uneven.

This chapter focuses on:

1. Matrix sparsity  
2. Library size variation  
3. Taxa prevalence  

## Load data

```{r load-phyloseq-object}
ps <- readRDS("data/moving-pictures-ps.rds")
```

## Plot theme

```{r cdi-plot-theme}
# ------------------------------------------------------------
# CDI Modern Plot Theme (Free Track Standard)
# ------------------------------------------------------------
cdi_teal <- "#036281"
cdi_yellow <- "#f7c546"
cdi_muted <- "#475569"

cdi_plot_theme <- function(base_size = 14) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        size = 22,
        face = "bold",
        hjust = 0.5
      ),
      plot.subtitle = ggplot2::element_text(
        size = 16,
        hjust = 0.5,
        margin = ggplot2::margin(b = 10)
      ),
      axis.title = ggplot2::element_text(
        face = "bold"
      ),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank()
    )
}
```

## Sparsity

```{r compute-sparsity}
otu <- methods::as(phyloseq::otu_table(ps), "matrix")
if (!phyloseq::taxa_are_rows(ps)) {
  otu <- t(otu)
}

zero_prop <- mean(otu == 0)
zero_prop
```

:::cdi-message
A high proportion of zeros is normal in microbiome data.

Filtering decisions are not cosmetic. They determine which patterns become visible.
:::

## Library size variation

```{r compute-library-size}
lib_size <- phyloseq::sample_sums(ps)
summary(lib_size)

df_lib <- data.frame(
  sample_id = names(lib_size),
  library_size = as.numeric(lib_size),
  stringsAsFactors = FALSE
)

dir.create("outputs/tables", recursive = TRUE, showWarnings = FALSE)
readr::write_csv(df_lib, "outputs/tables/library-size.csv")
```

### Library size distribution

```{r plot-library-size}
ggplot2::ggplot(df_lib, ggplot2::aes(x = library_size)) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 30,
    fill = cdi_teal,
    color = "white",
    alpha = 0.90
  ) +
  ggplot2::geom_density(color = cdi_yellow, linewidth = 1.2) +
  ggplot2::scale_y_continuous(labels = scales::label_number()) +
  ggplot2::labs(
    title = "Library size distribution",
    subtitle = "Sequencing depth varies across samples",
    x = "Library size (total reads per sample)",
    y = "Density"
  ) +
  cdi_plot_theme() +
  ggplot2::theme(
    plot.subtitle = ggplot2::element_text(color = cdi_muted)
  )
```

:::cdi-message
When sequencing depth varies, raw counts are not comparable.

Normalization is required for fair comparison across samples.
:::

## Detection and prevalence

```{r compute-prevalence}
detection <- 1

taxa_prev_n <- rowSums(otu >= detection)
taxa_prev_df <- data.frame(
  taxon = rownames(otu),
  prevalence_n = taxa_prev_n,
  prevalence_frac = taxa_prev_n / ncol(otu),
  stringsAsFactors = FALSE
)

summary(taxa_prev_df$prevalence_n)

dir.create("outputs/tables", recursive = TRUE, showWarnings = FALSE)
readr::write_csv(taxa_prev_df, "outputs/tables/taxa-prevalence.csv")
```

### Prevalence distribution

```{r plot-prevalence}
# Histogram view
ggplot2::ggplot(taxa_prev_df, ggplot2::aes(x = prevalence_n)) +
  ggplot2::geom_histogram(
    bins = 30,
    fill = cdi_teal,
    color = "white",
    alpha = 0.90
  ) +
  ggplot2::labs(
    title = "Taxon prevalence distribution",
    subtitle = "Most taxa are detected in few samples",
    x = "Number of samples detected",
    y = "Number of taxa"
  ) +
  cdi_plot_theme()

# ECDF view (often clearer for sparse data)
ggplot2::ggplot(taxa_prev_df, ggplot2::aes(x = prevalence_n)) +
  ggplot2::stat_ecdf(geom = "step", color = cdi_teal, linewidth = 1.2) +
  ggplot2::labs(
    title = "ECDF of taxon prevalence",
    subtitle = "Fraction of taxa detected by at least N samples",
    x = "Number of samples detected",
    y = "Fraction of taxa"
  ) +
  cdi_plot_theme()
```

:::cdi-message
Most taxa are observed in only a small fraction of samples.

Prevalence filtering improves clarity, but it also changes the feature space.
Filtering rules must be stated explicitly.
:::

## Simple prevalence filter

```{r prevalence-filter}
min_prev_frac <- 0.10

keep_taxa <- taxa_prev_df$taxon[taxa_prev_df$prevalence_frac >= min_prev_frac]
ps_prev <- phyloseq::prune_taxa(keep_taxa, ps)

c(original_taxa = phyloseq::ntaxa(ps), after_filter = phyloseq::ntaxa(ps_prev))
```

:::cdi-message
Prevalence filtering reduces dimensionality and improves readability.

It can also change ordination and diversity results. Interpretation should acknowledge these choices.
:::

## Relative abundance preview

```{r relative-abundance-preview}
ps_rel <- phyloseq::transform_sample_counts(ps, function(x) x / sum(x))
range(phyloseq::sample_sums(ps_rel))
```

:::cdi-message
After transformation, each sample sums to 1.

This enables clearer visualization, but does not remove compositional constraints.
:::

<!-- :::next
Next → [Composition Visualization](04-composition-visualization.qmd)
::: -->
